\section{Results and Discussion}

\subsection{NLP Component Results}

The NLP classifier was trained on 4,000 stratified samples for 3 epochs with the DistilBERT encoder frozen.

\begin{table}[h]
\centering
\caption{NLP Classifier Performance (800 validation samples)}
\label{tab:nlp_results}
\begin{tabular}{lcccc}
\toprule
\textbf{Class} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-Score} & \textbf{Support} \\
\midrule
Ham & 0.78 & 0.98 & 0.87 & 400 \\
Scam & 0.97 & 0.72 & 0.83 & 400 \\
\midrule
\textbf{Weighted Avg} & \textbf{0.87} & \textbf{0.85} & \textbf{0.85} & \textbf{800} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Overall Accuracy:} 85\%

\textbf{Analysis:}
\begin{itemize}
    \item High scam precision (0.97) indicates low false positive rate
    \item Ham recall (0.98) shows the model rarely misclassifies benign messages
    \item Scam recall (0.72) is acceptable but leaves room for improvement
    \item Training time: $\sim$5 minutes on CPU
\end{itemize}

\subsection{GNN Component Results}

The GNN was trained on the Elliptic dataset with focal loss and class weighting.

\begin{table}[h]
\centering
\caption{GNN Performance (9,313 validation nodes)}
\label{tab:gnn_results}
\begin{tabular}{lcccc}
\toprule
\textbf{Class} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-Score} & \textbf{Support} \\
\midrule
Licit & 0.98 & 0.79 & 0.87 & 8,840 \\
Illicit & 0.15 & 0.69 & 0.25 & 473 \\
\midrule
\textbf{Weighted Avg} & \textbf{0.94} & \textbf{0.78} & \textbf{0.84} & \textbf{9,313} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Overall Accuracy:} 78\%

\textbf{Analysis:}
\begin{itemize}
    \item Illicit recall (0.69) is the primary success metric—catches 69\% of fraud
    \item Low illicit precision (0.15) means high false positive rate, but acceptable because:
    \begin{itemize}
        \item False alarms are reviewed by analysts
        \item Fusion with NLP improves precision
        \item Missing fraud is more costly than false alarms
    \end{itemize}
    \item Training time: $\sim$3 minutes on CPU
\end{itemize}

\subsection{Fusion Model Results}

The fusion MLP was trained on 4,139 paired samples.

\begin{table}[h]
\centering
\caption{Fusion Model Performance (4,139 validation samples)}
\label{tab:fusion_results}
\begin{tabular}{lcccc}
\toprule
\textbf{Class} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-Score} & \textbf{Support} \\
\midrule
Benign & 0.99 & 0.98 & 0.99 & 3,230 \\
Scam & 0.94 & 0.97 & 0.95 & 909 \\
\midrule
\textbf{Weighted Avg} & \textbf{0.98} & \textbf{0.98} & \textbf{0.98} & \textbf{4,139} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Overall Accuracy:} 98\%

\subsection{Comparative Analysis}

\begin{table}[h]
\centering
\caption{Model Comparison}
\label{tab:comparison}
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Accuracy} & \textbf{Scam Prec.} & \textbf{Scam Recall} & \textbf{Scam F1} \\
\midrule
NLP Only & 85\% & 0.97 & 0.72 & 0.83 \\
GNN Only & 78\% & 0.15 & 0.69 & 0.25 \\
\textbf{Fusion} & \textbf{98\%} & \textbf{0.94} & \textbf{0.97} & \textbf{0.95} \\
\bottomrule
\end{tabular}
\end{table}

The fusion model achieves:
\begin{itemize}
    \item +13\% accuracy over NLP
    \item +20\% accuracy over GNN
    \item +25\% scam recall over NLP
    \item +79\% scam precision over GNN
\end{itemize}

\subsection{RAG Module Results}

For the test message "Your SBI account will be blocked. Update KYC now.", RAG retrieved:

\begin{table}[h]
\centering
\caption{RAG Retrieval Results}
\label{tab:rag_results}
\begin{tabular}{clcc}
\toprule
\textbf{Rank} & \textbf{Pattern} & \textbf{Label} & \textbf{Distance} \\
\midrule
1 & "Your KYC is expired..." & kyc\_scam & 0.386 \\
2 & "SBI account is suspended..." & phishing & 0.503 \\
3 & "UPI PIN compromised..." & phishing & 1.099 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Analysis:}
\begin{itemize}
    \item Top match (distance 0.386) correctly identifies KYC scam pattern
    \item Cosine distance $<$0.5 indicates high similarity
    \item Average retrieval time: 50ms
\end{itemize}

\subsection{System Performance}

\begin{table}[h]
\centering
\caption{Inference Latency Breakdown}
\label{tab:latency}
\begin{tabular}{lc}
\toprule
\textbf{Component} & \textbf{Latency (ms)} \\
\midrule
Text preprocessing & 5 \\
DistilBERT embedding & 180 \\
NLP classification & 15 \\
LLM intent (fallback) & 10 \\
RAG retrieval & 50 \\
GNN encoding & 45 \\
Fusion MLP & 8 \\
\midrule
\textbf{Total (without LLM)} & \textbf{313} \\
\textbf{Total (with LLM)} & \textbf{15,000-20,000} \\
\bottomrule
\end{tabular}
\end{table}

The system meets real-time requirements ($<$2s) when LLM is disabled or cached.

\subsection{Comparison with Baseline Methods}

\begin{table}[h]
\centering
\caption{Comparison with Baseline Approaches}
\label{tab:baselines}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{Accuracy} & \textbf{Scam F1} & \textbf{Explainable} & \textbf{Real-time} \\
\midrule
Rule-based & 62\% & 0.58 & \checkmark & \checkmark \\
Logistic Regression & 78\% & 0.71 & $\times$ & \checkmark \\
BERT (text only) & 89\% & 0.85 & $\times$ & $\times$ \\
GCN (graph only) & 76\% & 0.23 & $\times$ & \checkmark \\
\textbf{Our Fusion} & \textbf{98\%} & \textbf{0.95} & \checkmark & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

Our approach outperforms all baselines while providing explainability and real-time inference.

\subsection{Limitations}

\begin{enumerate}
    \item \textbf{LLM Latency:} Mistral 7B is too slow for real-time use on CPU
    \item \textbf{Dataset Pairing:} Synthetic pairing may not reflect real-world correlations
    \item \textbf{Graph Scalability:} GAT has $O(N^2)$ complexity for dense graphs
    \item \textbf{Cold Start:} New users without transaction history cannot benefit from GNN
    \item \textbf{Language:} Currently English-only
    \item \textbf{Adversarial Robustness:} Not evaluated against adversarial attacks
\end{enumerate}

\subsection{Detailed Error Analysis}

\subsubsection{False Negative Analysis}

We manually analyzed 50 false negatives (scams classified as safe) to identify failure patterns:

\textbf{Category 1: Subtle Phishing (32\%)}
\begin{itemize}
    \item Messages without obvious urgency keywords
    \item Example: "Dear customer, please verify your account details at your convenience"
    \item Root cause: Lack of explicit threat language
\end{itemize}

\textbf{Category 2: Novel Tactics (28\%)}
\begin{itemize}
    \item Scam types not present in training data
    \item Example: Cryptocurrency investment scams, NFT fraud
    \item Root cause: Training data bias toward traditional banking scams
\end{itemize}

\textbf{Category 3: Code-Switching (24\%)}
\begin{itemize}
    \item Messages mixing English with Hindi/regional languages
    \item Example: "Aapka account block ho jayega. Update KYC now"
    \item Root cause: Monolingual training data
\end{itemize}

\textbf{Category 4: Legitimate Urgency (16\%)}
\begin{itemize}
    \item Genuine urgent notifications misclassified as scams
    \item Example: "Your credit card payment is due today. Pay now to avoid late fees"
    \item Root cause: Difficulty distinguishing legitimate urgency from manipulation
\end{itemize}

\subsubsection{False Positive Analysis}

Analysis of 30 false positives (safe messages classified as scams):

\textbf{Category 1: Marketing Messages (43\%)}
\begin{itemize}
    \item Promotional content with urgency language
    \item Example: "Limited time offer! Claim your cashback before midnight"
    \item Root cause: Overlap between marketing and scam language patterns
\end{itemize}

\textbf{Category 2: Legitimate Alerts (37\%)}
\begin{itemize}
    \item Genuine security alerts from banks
    \item Example: "Unusual activity detected on your account. Please verify"
    \item Root cause: Similar phrasing to phishing attempts
\end{itemize}

\textbf{Category 3: Transaction Notifications (20\%)}
\begin{itemize}
    \item Payment confirmations with large amounts
    \item Example: "You have received Rs 49,999 from XYZ"
    \item Root cause: Amount-based heuristics triggering false alarms
\end{itemize}

\subsection{Case Studies}

\subsubsection{Case Study 1: KYC Phishing Detection}

\textbf{Input Message:}
\begin{quote}
"URGENT: Your PAN-Aadhaar linking will expire in 24 hours. Update KYC immediately at [URL] or your bank account will be frozen. Call 9876543210 for assistance."
\end{quote}

\textbf{System Output:}
\begin{itemize}
    \item Risk Score: 0.96
    \item Label: SCAM
    \item Detected Intent: phishing
    \item Tactics: [urgency, authority, impersonation]
    \item Entities: [PAN, Aadhaar, bank, phone number]
    \item RAG Match: "Your KYC is expired..." (distance: 0.31)
\end{itemize}

\textbf{Analysis:}
The system correctly identified multiple red flags: artificial urgency (24 hours), authority impersonation (government documents), and suspicious contact information. The RAG module matched it to a known KYC scam pattern with high confidence.

\subsubsection{Case Study 2: Lottery Scam with Transaction Pattern}

\textbf{Input Message:}
\begin{quote}
"Congratulations! You have won Rs 25 lakh in the Kaun Banega Crorepati lucky draw. Send Rs 5,000 processing fee to claim your prize."
\end{quote}

\textbf{Transaction Features:}
\begin{itemize}
    \item Amount: Rs 5,000
    \item Time: 2:00 AM
    \item Day: Sunday
    \item New account: Yes
\end{itemize}

\textbf{System Output:}
\begin{itemize}
    \item Risk Score: 0.98
    \item Label: SCAM
    \item NLP Contribution: 0.71
    \item GNN Contribution: 0.27
\end{itemize}

\textbf{Analysis:}
Both modalities contributed to detection. NLP identified lottery scam language, while GNN flagged suspicious transaction timing (2 AM) and new account status. The fusion model combined these signals for high-confidence prediction.

\subsubsection{Case Study 3: Legitimate Urgent Notification}

\textbf{Input Message:}
\begin{quote}
"Your HDFC credit card payment of Rs 15,230 is due today. Pay now to avoid late payment charges. Visit HDFC NetBanking or call 1800-XXX-XXXX."
\end{quote}

\textbf{System Output:}
\begin{itemize}
    \item Risk Score: 0.42
    \item Label: SAFE
    \item Detected Intent: benign
    \item Tactics: [urgency]
\end{itemize}

\textbf{Analysis:}
Despite urgency language, the system correctly classified this as safe because: (1) official HDFC toll-free number, (2) specific payment amount, (3) legitimate banking terminology, (4) no suspicious URL or phone number patterns.

\subsection{Ablation Study}

To understand component contributions, we performed an ablation study:

\begin{table}[h]
\centering
\caption{Ablation Study Results}
\label{tab:ablation}
\begin{tabular}{lcc}
\toprule
\textbf{Configuration} & \textbf{Accuracy} & \textbf{Scam F1} \\
\midrule
Full Model & 98\% & 0.95 \\
Without LLM Intent & 97\% & 0.94 \\
Without RAG & 96\% & 0.93 \\
Without GNN & 85\% & 0.83 \\
Without NLP & 78\% & 0.25 \\
Without Explainability & 98\% & 0.95 \\
\midrule
Random Baseline & 50\% & 0.33 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Key Findings:}
\begin{itemize}
    \item NLP is the most critical component (13\% accuracy drop without it)
    \item GNN provides significant lift (13\% accuracy gain)
    \item LLM and RAG provide marginal improvements (1-2\%)
    \item Explainability does not affect prediction accuracy (as expected)
\end{itemize}

\subsection{Computational Efficiency Analysis}

\begin{table}[h]
\centering
\caption{Model Size and Inference Time}
\label{tab:efficiency}
\begin{tabular}{lccc}
\toprule
\textbf{Component} & \textbf{Parameters} & \textbf{Size (MB)} & \textbf{Inference (ms)} \\
\midrule
DistilBERT (frozen) & 66M & 268 & 180 \\
NLP Classifier Head & 99K & 0.4 & 15 \\
GNN (GAT) & 1.2M & 4.8 & 45 \\
Fusion MLP & 108K & 0.4 & 8 \\
\midrule
\textbf{Total} & \textbf{67.4M} & \textbf{273.6} & \textbf{248} \\
\bottomrule
\end{tabular}
\end{table}

The system is lightweight enough for CPU deployment, with total model size under 300 MB and inference time under 300ms (excluding LLM).
