\documentclass[12pt,a4paper]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{listings}
\usepackage{xcolor}

% Spacing
\onehalfspacing

% Code listing style
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    language=Python,
    showstringspaces=false,
    commentstyle=\color{gray},
    keywordstyle=\color{blue},
    stringstyle=\color{red}
}

% Title
\title{\textbf{Early Detection of Financial Scams and Fraudulent Messages Using Context-Aware NLP and Graph-Based Learning}}

\author{
    Malhar Shinde \\
    \textit{Symbiosis Institute of Technology, Pune Campus} \\
    \textit{Symbiosis International (Deemed University)} \\
    \textit{Pune, India} \\
    \texttt{malhar.shinde@sitpune.edu.in}
}

\date{April 25, 2026}

\begin{document}

\maketitle

\begin{abstract}
Financial fraud and scam messages have become more and more sophisticated and caused huge economic losses worldwide. Traditional rule-based detection systems are not able to adapt to the changing fraud patterns and are not able to analyse textual content and transactional relationships simultaneously. In this work, we propose a novel multi-modal framework with Natural Language Processing (NLP) and Graph Neural Networks (GNN) for real-time detection of financial scams. The system utilises DistilBERT for semantic text analysis, Graph Attention Networks (GAT) for transaction pattern identification, and a Large Language Model (LLM) for intent extraction. A Retrieval Augmented Generation (RAG) component supplies pattern memory by matching incoming messages to known scam databases. The fused model aggregates the embeddings from both modalities to achieve 98\% accuracy on paired text-transaction data. The individual components achieved 85\% accuracy on the NLP task and 69\% recall on the GNN task. The system is also explainable through SHAP analysis and attention weight visualisation, making it suitable for regulatory compliance. The framework is deployed as a FastAPI service with Streamlit interface, showing its practical applicability for real-time fraud detection for banking and payment systems.
\end{abstract}

\noindent\textbf{Keywords:} Financial Fraud Detection, Natural Language Processing, Graph Neural Networks, Multi-Modal Learning, Explainable AI, Retrieval-Augmented Generation

\newpage
\tableofcontents
\newpage

\section{Introduction}

\subsection{Background}

The proliferation of digital payment systems and online banking has revolutionized financial transactions, but it has also created new opportunities for fraudulent activities. Financial scams manifest in multiple forms: phishing emails impersonating banks, SMS messages creating urgency around KYC updates, lottery scams, and sophisticated social engineering attacks. According to the Reserve Bank of India, digital payment fraud cases increased by 300\% between 2020 and 2023, with losses exceeding ₹1,200 crores annually \cite{rbi2023}.

Traditional fraud detection systems rely on rule-based approaches or simple machine learning classifiers that analyze either textual content or transaction patterns in isolation. However, modern scams exploit both channels simultaneously—a fraudulent message may accompany a suspicious transaction pattern, and neither signal alone provides sufficient evidence for detection.

\subsection{Motivation}

The limitations of existing approaches motivated this research:

\begin{enumerate}
    \item \textbf{Unimodal Analysis:} Most systems analyze text or transactions separately, missing correlations between message content and behavioral patterns.
    \item \textbf{Static Pattern Matching:} Rule-based systems cannot adapt to novel scam tactics without manual updates.
    \item \textbf{Lack of Explainability:} Black-box models provide predictions without justification, making them unsuitable for regulatory compliance.
    \item \textbf{Cold Start Problem:} New scam patterns require extensive labeled data before detection becomes effective.
    \item \textbf{Real-Time Requirements:} Banking systems need sub-second response times.
\end{enumerate}

\subsection{Problem Statement}

Given a financial message (SMS, email, or payment note) and associated transaction metadata, the system must:

\begin{itemize}
    \item Classify the message as SCAM or SAFE with high precision and recall
    \item Identify manipulation tactics and extracted entities
    \item Detect anomalous transaction patterns in the user's behavioral graph
    \item Provide human-interpretable explanations for predictions
    \item Operate in real-time ($<$2 seconds per prediction)
    \item Adapt to new scam patterns without full model retraining
\end{itemize}

\subsection{Objectives}

The primary objectives of this project are:

\begin{enumerate}
    \item \textbf{Multi-Modal Integration:} Develop a framework that jointly analyzes textual and graph-structured data.
    \item \textbf{High Performance:} Achieve $>$90\% accuracy while maintaining $>$85\% recall.
    \item \textbf{Explainability:} Provide feature attribution, attention visualization, and natural language explanations.
    \item \textbf{Adaptability:} Implement RAG-based pattern memory for zero-shot detection.
    \item \textbf{Production Readiness:} Deploy as a scalable API with monitoring and compliance features.
\end{enumerate}

\subsection{Contributions}

This work makes the following contributions:

\begin{enumerate}
    \item \textbf{Novel Architecture:} First framework to combine DistilBERT, GAT, and LLM-based intent extraction in a unified pipeline for financial fraud detection.
    \item \textbf{Late Fusion Strategy:} Demonstrates that training NLP and GNN components independently before fusion improves modularity.
    \item \textbf{RAG Integration:} Shows that retrieval-augmented generation enables zero-shot adaptation to new scam patterns.
    \item \textbf{Explainability Framework:} Combines SHAP, attention weights, and LLM-generated reports for multi-level interpretability.
    \item \textbf{Open-Source Implementation:} Provides complete, reproducible codebase with training scripts, API, and UI.
\end{enumerate}

\subsection{Report Organization}

The remainder of this report is organized as follows: Section 2 reviews related work in fraud detection, NLP, and GNN applications. Section 3 describes the methodology, including dataset preparation, model architectures, and training procedures. Section 4 presents experimental results and comparative analysis. Section 5 discusses findings, limitations, and implications. Section 6 concludes with future research directions.

\input{sections/literature_review}
\input{sections/methodology}
\input{sections/results}
\input{sections/conclusion}

\bibliographystyle{IEEEtran}
\bibliography{references}

\appendix
\input{sections/appendix}

\end{document}
